### 2012-08-15 paper/chemokine_main.tex
Current Draft after Steph Edits

@@

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci representing a tiny fraction of the total lung.  T-cell navigation through the complex branching bronchial vascular network is enabled by local cytokine and chemokine production from infected epithelial cells.  Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.  Of the eight chemokines tested only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (results only) with IP-10 always dominating the affected T-cell migration direction.  A delayed differential equation model was fit to the empirical data, revealing per cell chemokine production rates for each strain.  These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects of inter-strain variation of chemokine production on T-cell recruitment.  In the model immune response is unable to clear the infection of pandemic H1N1 influenza due to the high rate of viral production and plaque size expansion.  The spatial nature of the model reveals unique challenges to T-cell recruitment and motility not visible in spatially homogeneous ODE models.  Expanding plaque sizes isolate infected cells, causing T-cell discovery and subsequent clearance of the infection to become less effective over time.

@@

Here we focus on the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).  To develop a model with tractable simulations, some key components of host defense are not incorporated explicitly and the model therefore does not aim to predict the true outcome for each strain examined.   We ask how do very small foci of infected tissue attract limited numbers of activated CD8 T cells after release into the systemic vascular system.  The circulating blood supply services non-infected tissue orders of magnitude larger than the target infected tissue, necessitating an efficient mechanism to focus cell migration.   A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Miao2010, Zhao2000, LiJeon2002}.  While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.

To characterize the local chemokine-secretion environment in which T cells migrate, we used data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.   We selected in vitro chemokine data as a tissue environment surrogate, because it may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and it allowed us to represent sustained chemokine gradients within the model lung.  Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to populate the ABM model with activated CD8+ T cells. [REF]

The standard hypothesis would posit that in this model the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.  While we found this to be true, we observed three interesting characteristics of chemokine-directed migration not revealed from experimental data.  First, of the only two chemokines stimulated by all three strains, the influence of IP-10 consistently dominated the effect of RANTES, regardless of strain.  Second, each chemokine exhibited a clear threshold to direct migration, with lesser impact at increasing concentrations above the threshold.  Finally, when the target (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient, possibly contributing to the loss of control of the pandemic strain.

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=0.5\textwidth]{SystemChart}
\end{center}
\caption{A simplified model of T cell search.  Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through the 14 layers of the branching bronchial network.  Upon reaching a capillary, T cells will exit into the tissue in the presence of a cytokine signal.  T cells that do not sense a signal will continue to recirculate either through the lymph network or through the pulmonary vein back to the top of the network.}
\label{fig:systemchart}
\end{figure}

Computational modeling was conducted using CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response. A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions. To model a growing infection, the lung was represented as a two dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node. Activated T cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection. T cells that do not encounter cytokine recirculate to the lymph node. At the site of infection when a T cell encounters an infected epithelial cell it induces apoptosis.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine. Virus diffuses locally, infecting nearby cells, and continuing the cycle. Chemokine also diffuses from secreting cells, creating a ball of stimulation around the infected region. After a delay to simulate lymph node stimulation and T cell proliferation, activated T cells begin exiting the lymph node and circulating through the vasculature to the tissue. Because T  cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations.

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{ModelChart}
\end{center}
\caption{A visual representation of the model.  Healthy epithelial cells infected by virus begin secreting virus after an incubation delay.  T cells are activated in the lymph node and traverse the bronchial vascular network to the site of infection where they are recruited by inflammatory cytokine.  Chemotaxing T cells climb the chemokine gradient and induce apoptosis in infected cells.  Solid arrows represent the transition of a cell from one behavioral state to another.  Dashed arrows display the mechanism used to induce a transition.  Dotted arrows show the production of new virus.}
\label{fig:modelchart}
\end{figure}

Epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}. \emph{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from {incubating} to \emph{expressing}. \emph{Expressing} cells secrete virus and chemokine for 17 hours and then die. \emph{Expressing} cells become \emph{apoptotic} if they encounter activated T-cells. Apoptoic cells continue to secrete virus until they die one hour after their transition. \emph{Dead} cells remain inert and do not regenerate over the course of an infection.

T-cells have three states: \emph{searching}, \emph{recirculating}, and \emph{chemotaxing}. \emph{Searching} T-Cells begin emerging from the lymph node at five days post infection. These T-Cells arrive at a random location on the lung's surface, perform a random walk in the tissue for 10 minutes and transition to \emph{circulating} in the absence of chemokine. \emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{searching} and are reintroduced to a new random location in the lung. When a \emph{searching} T-Cell encounters chemokine, it converts to \emph{chemotaxing} and begins climbing the chemotactic gradient to the source of infection. \emph{Chemotaxing} T-Cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells. \emph{Chemotaxing} decay exponentially with an average lifespan of two hours.

@@

\begin{table}
\begin{center}
\begin{tabular}{ | c | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
  Viral Decay &  $1/day$ & Phenomenological \\
  Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
  Chemokine Decay &  $3.8508$e-4$/s$ & 30m half-life \\
  Infection Sensitivity &  $2 h/virion$ & Phenomenological \\
  Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
  Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
  Apoptosis Time & $1 h$ & Phenomenological \\
  T-Cell Production Rate & $777/h$ & \cite{Miao2010} \\ 
  Circulation Time & $6 m$ & \cite{Banerjee2010b} \\
  Search Time & $10 m$ & \cite{Banerjee2010b} \\
  T-Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
  T-Cell Speed (Activated) & $3 \mu m/s$ & \cite{Miller2003} \\
  T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
  T-Cell Expected Kill Rate & $10 m$ & [REF] \\
  Cell Radius & $25 \mu m$ & Phenomenological \\
  T-Cell Age (at FOI) & $2 h$ & [REF] \\
  T-Cell Age (Other) & $3 d$ & [REF] \\
  T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
  IgM Strength & Viral decay of $3/day$ & Phenomenological \\
  \hline  
\end{tabular}
\caption{Generic model parameters}
\label{table:parameters}
\end{center}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{ | r | c | c | c | }
  \hline                        
  Parameter & Value & 95\% Confidence & Source \\
   & \footnotesize{(pg/s$\cdot$cell)}  & \footnotesize{(pg/s$\cdot$cell)} &  \\
  \hline
  \textbf{IP-10 Production Rate} & & & ODE Fits \\
  Avian H5N1 & 2.0e-4 & \small{8.4e-5 --- 4.2e-4} & \\
  Seasonal H1N1 & 1.8e-4 & \small{1.2e-4 --- 3.0e-4} & \\
  Pandemic H1N1 & 8.7e-5 & \small{1.7e-5 --- 7.1e-4}  & \\
  \hline
  \textbf{RANTES Production Rate} & & & ODE Fits \\
  Avian H5N1 & 1.3e-5 & \small{7.9e-6 --- 1.9e-5} & \\
  Seasonal H1N1 & 8.9e-7 & \small{4.8e-7 --- 1.6e-6} & \\
  Pandemic H1N1 & 4.4e-6 & \small{5.0e-7 --- 3.6e-5} & \\
%  \hline
  \textbf{Virus Production Rate} & & & \cite{Mitchell2011} \\
  Avian H5N1 & 5.4e-5 PFU/s$\cdot$cell & & \\
  Seasonal H1N1 & 3.8e-4 PFU/s$\cdot$cell & & \\
  Pandemic H1N1 & 5.1e-3 PFU/s$\cdot$cell & & \\  
  \hline  
\end{tabular}
\caption{Strain specific model parameters}
\label{table:strains}
\end{center}
\end{table}

@@

Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).  IP-10 concentration increases were observed by 8h p.i., and RANTES by 16h p.i..  To estimate per-cell production rates, we extended the ODE model of Ref. \cite{Mitchell2011} to represent chemokine production from infected cells.  Model fits, shown in Table \ref{table:strains}, were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{Mitchell2011}.  These expression rates are similar across different strains with the exception of significantly higher RANTES production in aH5N1.  There is no discernible correlation between viral production and induced chemokine production across the three strains.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{data}
 \end{center}
\caption{Empirical data of viral and cytokine titer for three strains of influenza: Avian H5N1 (A), Seasonal sH1N1 (B), and Pandemic pH1N1 (C).  For each strain, the viral load (blue circle) is shown in PFU/mL and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.  sH1N1 IP-10 registered three values not included in model fitting above the measurement threshold of 8,500 $pg/mL$ (empty red triangles).  Solid lines show model fits to IP-10 and RANTES data.} 
 \label{fig:data}
\end{figure}

@@

T-Cells exit a capillary when they sense inflammatory cytokine and follow the chemokine gradient.  The model uses chemokine gradient as a surrogate for the inflammatory response.  Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.  T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.

To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude. We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on infection kinetics.  Because multiple T-cells in the same area do not increase apoptosis, we hypothesize that above a critical threshold no additional benefit is conferred by increasing T cell presence.  We use the sensitivity of 10 $ng/mL$ for each run, which corresponds to a $1 nM$ concentration assuming a chemokine molecular weight of 10 $kDa$ \cite{Gao2003}.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{sensitivity}
 \end{center}
\caption{Varying T-cell sensitivity to chemokine in the model: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES. Total number of infected, expressing and apoptotic cells are plotted for each infection.  The sensitivity value specifies the minimum level of chemokine concentration required for T-cells to detect it. } 
 \label{fig:sensitivity}
\end{figure}

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T-cells to the infected cells.  Although virus is produced at a higher rate than chemokine, the larger particle diffuses much more slowly.  Chemokine decays more quickly.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).

Up to day 4 the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare. Over time, cells in the plaque's interior die, and active cells are a decreasing proportion of the plaque. T-cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  In aH5N1, the plaque is dense, allowing T-cells to find secreting cells easily and infection is eliminated.  However, in both H1N1 simulations secreting cells were not eliminated.  Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..  T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.  Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T-cells can lag behind cellular changes in the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T-cells from completely clearing the infection.

The immune response fails to clear sH1N1 completely and is unable to contain pH1N1 (Fig. \ref{fig:sensitivity}).  T-cells are unable to kill cells that have not yet presented antigen.  At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}). By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.  The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T-cells can control (but not eliminate) the sH1N1 infection.

A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.  That is, a new viral particle is produced approximately every 200 seconds.  Assuming that secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do is to limit production to 18 new viral particles.  T-Cells alone cannot contain the pH1N1 infection.  In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T-cells to limit a single infected cell to 1.3 viral particles in the one-hour window.  Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{cycells}
 \end{center}
\caption{A sH1N1 infection in CyCells. Screenshots taken at Day 4, Day 5.5, and Day 7.  The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells introduced at day 5 (green).  Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL (see the legend in the bottom right).} 
 \label{fig:cycells}
\end{figure}

@@

@@

Hypercytokinemia is a characteristic of virulent influenza strains \cite{DeJong2006} but bloodstream levels of cytokines and chemokines may not reflect local concentrations and dynamics critical in guiding T cells to infected epithelial cells.  We elected to use our data on in vitro secretion of chemokines by human bronchial epithelial (NHBE) cells obtained from the same cultures of the three virus strains previously reported \cite{Mitchell2011}.  In this study the replication kinetics and productivity of the three virus strains were markedly different, but monolayer infection was initiated with a low MOI of 0.01 or 0.001, mimicking the initial infection naturally.  The levels of the chemokines CCL5 and CXCL10 in the basal media induced by the 3 strains were approximately equivalent at 24h post-infection.  When the productivity per cell was calculated, the avian H5N1 strain induced higher secretion of CCL5 (RANTES) than the seasonal or pandemic strains.  Greater CCL5 secretion by H5N1 strains has been noted in in vitro NHBE cultures when a relatively high MOI is used compared to seasonal strains \cite{Chan2005, Chan2010, Zeng2011} and pandemic strains \cite{Zeng2011}.  MOI and degree of differentiation of NHBE are critical determinants of cyto/chemokine secretion \cite{Chan2010}.  However, in spite of the attenuated type 1 interferon response induced by H5N1 strains \cite{Zeng2007}, the chemokine secretion is not attenuated.

>>>>

\usepackage{multirow}
\renewcommand{\arraystretch}{1.1}

@@

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci comprising a tiny fraction of the total lung.  T-cell navigation to the infected region is enabled by local cytokine and chemokine production from infected epithelial cells.  Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.  Of the eight chemokines tested, only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (results only) with IP-10 always dominating the affected T-cell migration direction.  A delayed differential equation model was fit to the empirical data, revealing per cell chemokine production rates for each strain.  These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects of inter-strain variation of chemokine production on T-cell recruitment and infection control.  T cells appear to be sufficiently sensitive to chemokine to function as well as they possibly can.  The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production and large plaque size.  The spatial nature of the model reveals unique challenges to T-cell recruitment and motility not visible in spatially homogeneous ODE models.  Expanding plaque sizes isolate infected cells, impeding T-cell discovery and subsequent clearance of the infection over time.

@@

Here we study the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).  To develop a model with tractable simulations, some components of host defense are abstracted and the model therefore does not aim to predict the true outcome for each strain examined.   We ask how very small foci of infected tissue can attract limited numbers of activated CD8 T cells after release into the systemic vascular system.  The circulating blood supply services non-infected tissue orders of magnitude larger than the target infected tissue, necessitating an efficient mechanism to focus cell migration.   A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Miao2010, Zhao2000, LiJeon2002}.  While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.

To characterize the local chemokine-secretion environment in which T cells migrate, we used data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.   We selected in vitro chemokine data as a tissue environment surrogate, because it may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and it allowed us to represent sustained chemokine gradients within the model lung.  Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to calibrate the ABM model. [REF]

The standard hypothesis posits that the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.  Our model confirms this hypothesis, but we also observed three interesting characteristics of chemokine-directed migration not revealed by earlier data.  First, of the only two chemokines stimulated by all three strains, the influence of IP-10 consistently dominated the effect of RANTES, regardless of strain.  Second, each chemokine exhibited a clear threshold to direct migration, with lesser impact at increasing concentrations above the threshold.  Finally, when the target (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient, possibly contributing to the loss of control of the pandemic strain.

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=0.5\textwidth]{SystemChart}
\end{center}
\caption{Model of T cell search.  Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through the 14 layers of the branching bronchial network.  Upon reaching a capillary, T cells exit into tissue if cytokine signal is present.  In the absence of signal, the T cell recirculates either through the lymph network or through the pulmonary vein back to the top of the network.}
\label{fig:systemchart}
\end{figure}

Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response. A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions. The lung was represented as a two-dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node. Activated T cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection. T cells that do not encounter cytokine recirculate to the lymph node. At the site of infection when a T cell encounters an infected epithelial cell it induces apoptosis.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine. Virus diffuses locally, infecting nearby cells, and continuing the cycle. Chemokine diffuses from secreting cells, creating a ball of stimulation around the infected region. After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells exit the lymph node and circulate through the vasculature to the tissue. Because T  cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations.

@@

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{ModelChart}
\end{center}
\caption{Visual representation of the model.  Healthy epithelial cells infected by virus begin secreting virus after the incubation delay.  Activated T cells traverse the bronchial vascular network and may be recruited by inflammatory cytokine.  Chemotaxing T cells climb the chemokine gradient and induce apoptosis in infected cells.  Solid arrows represent a cell state transition from one behavior to another.  Dashed arrows display the mechanism used to induce a transition.  Dotted arrows indicate the production of new virus.}
\label{fig:modelchart}
\end{figure}

In the model, epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}. \emph{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from {incubating} to \emph{expressing}. \emph{Expressing} cells secrete virus and chemokine for 17 hours and then die. \emph{Expressing} cells become \emph{apoptotic} if they encounter activated T-cells. Apoptoic cells continue to secrete virus until they die one hour after their transition. \emph{Dead} cells remain inert and do not regenerate over the course of an infection.

T-cells have three states. \emph{Searching} T-Cells begin emerging from the lymph node at five days post infection. \textit{Searching} T-Cells arrive at a random location on the lung's surface, wander randomly in the tissue for 10 minutes, and transition to \emph{circulating} in the absence of chemokine. \emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{searching} and are reintroduced to a new random location in the lung. When a \emph{searching} T-Cell encounters chemokine, it converts to \emph{chemotaxing} and begins following the chemotactic gradient to the source of infection. \emph{Chemotaxing} T-Cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells. \emph{Chemotaxing} cells decay exponentially with an average lifespan of two hours.

@@

\begin{table}
\begin{center}
\begin{tabular}{ | c | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
  Viral Decay &  $1/day$ & Phenomenological \\
  Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
  Chemokine Decay &  $3.8508$e-4$/s$ & 30m half-life \\
  Infection Sensitivity &  $2 h/virion$ & Phenomenological \\
  Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
  Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
  Apoptosis Time & $1 h$ & Phenomenological \\
  T-Cell Production Rate & $777/h$ & \cite{Miao2010} \\ 
  Circulation Time & $6 m$ & \cite{Banerjee2010b} \\
  Search Time & $10 m$ & \cite{Banerjee2010b} \\
  T-Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
  T-Cell Speed (Activated) & $3 \mu m/s$ & \cite{Miller2003} \\
  T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
  T-Cell Expected Kill Rate & $10 m$ & [REF] \\
  Cell Radius & $25 \mu m$ & Phenomenological \\
  T-Cell Age (at FOI) & $2 h$ & [REF] \\
  T-Cell Age (Other) & $3 d$ & [REF] \\
  T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
  IgM Strength & Viral decay of $3/day$ & Phenomenological \\
  \hline  
\end{tabular}
\caption{Model parameters}
\label{table:parameters}
\end{center}
\end{table}

\begin{table}
\centering
\begin{tabular}{ | r | c | c | c | }
  \hline                        
  \multicolumn{1}{|c|}{\multirow{2}{*}{Strain}} & IP-10 Production & RANTES Production & Viral Production \\
   & \footnotesize{(pg/s$\cdot$cell)}  & \footnotesize{(pg/s$\cdot$cell)} &  \footnotesize{(PFU/s$\cdot$cell)} \\
  \hline
  \multirow{2}{*}{Avian H5N1} & 2.0e-4 &  1.3e-5 & \multirow{2}{*}{5.4e-5} \\
   &  \footnotesize{8.4e-5 --- 4.2e-4} & \footnotesize{7.9e-6 --- 1.9e-5} & \\ 
   \hline
  \multirow{2}{*}{Seasonal H1N1} & 1.8e-4 &  8.9e-7 & \multirow{2}{*}{3.8e-4} \\
   & \footnotesize{1.2e-4 --- 3.0e-4} & \footnotesize{4.8e-7 --- 1.6e-6} &  \\
   \hline
  \multirow{2}{*}{Pandemic H1N1} & 8.7e-5 &  4.3e-6 & \multirow{2}{*}{5.1e-3} \\
   & \footnotesize{1.7e-5 --- 7.1e-4} & \footnotesize{5.0e-7 --- 3.5e-5} &  \\
  \hline
\end{tabular}
\caption{Strain-specific model parameters.  Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \cite{Wu1986}.  Viral production values are taken from \cite{Mitchell2011}.}
\label{table:strains}
\end{table}

@@

Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..  To estimate per-cell production rates, we extended the ODE model of Ref. \cite{Mitchell2011} to represent chemokine production from infected cells (Supplemental Methods).  Model fits, (Table \ref{table:strains}), were computed for three strains (Fig.~\ref{fig:data}) using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).  Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.  There is no positive correlation between viral production and induced chemokine production across the three strains.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{data}
 \end{center}
\caption{Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.  Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.  sH1N1 IP-10 registered three values, not included in model fitting, above the measurement threshold of 8,500 $pg/mL$ (empty red triangles).  Dashed lines show model fits to IP-10 and RANTES data.} 
 \label{fig:data}
\end{figure}

@@

The model uses chemokine gradient as a surrogate for the inflammatory cytokines required for extravasation.  Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.  However, T-cell sensitivity determines how strong the aggregate chemokine signal must be.

We simulated T cell sensitivity levels ranging over seven orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) between concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on T cell movement.  Because multiple T-cells in the same area do not enhance control of infection, we hypothesize more T cells are not better above a critical threshold.  We set the sensitivity to 10 $ng/mL$ (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.  Single runs were used due to computational limitations.  Model variance is discussed in the Supplemental Results.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{sensitivity}
 \end{center}
\caption{Varying T-cell sensitivity to chemokine: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES. Total number of infected, expressing and apoptotic cells are plotted for each infection.  The sensitivity value specifies the minimum level of chemokine concentration required for T-cells to detect it. } 
 \label{fig:sensitivity}
\end{figure}

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T-cells to the infected cells.  Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}). Until day 4 the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare. Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque. T-cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  In aH5N1, the plaque is dense, allowing T-cells to find secreting cells easily, and infection is eliminated.  However, in both H1N1 simulations secreting cells were not eliminated.  Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..  T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.  Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T-cells can fail to detect cellular changes in the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T-cells from completely clearing the infection.

T-cells are unable to kill cells that have not yet presented antigen.  At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}). By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.  The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T-cells can control (but not eliminate) the sH1N1 infection.

A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.  That is, in each secreting cell a new viral particle is produced approximately every 200 seconds.  Assuming that secretion continues for one hour after apoptosis is initiated, the best a T-Cell could do is limit production to 18 new viral particles.  Thus, T-Cells alone cannot contain the pH1N1 infection.  In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T-cells to limit a single infected cell to 1.3 viral particles in the one-hour window.  Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{cycells}
 \end{center}
\caption{Simulated sH1N1 infection. Screenshots from day 4, day 5.5, and day 7.  The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells arriving at day 5 (green).  Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL (see the legend in the bottom right).} 
 \label{fig:cycells}
\end{figure}

@@

\subsection*{Caveats}

Parameters are estimated to the best of our knowledge when values could not be found in literature.  We choose to ignore most aspects of the innate immune response to avoid over-parameterizing the model.  The unbounded growth of pH1N1 can partially be attributed to the lack of many innate mechanisms.  T cells are assumed to have the same level of sensitivity to both IP-10 and RANTES.  Furthermore, T cells climb chemokine gradient in straight lines as our experiments with biased random walks showed no effect on model behavior.

@@

Hypercytokinemia is a characteristic of virulent influenza strains \cite{DeJong2006}, but bloodstream levels of cytokines and chemokines may not reflect local concentrations or dynamics critical in guiding T cells to infected epithelial cells.  We elected to use our data on in vitro secretion of chemokines by human bronchial epithelial (NHBE) cells obtained from the same cultures of the three virus strains previously reported \cite{Mitchell2011}.  In this study the replication kinetics and productivity of the three virus strains were markedly different, but monolayer infection was initiated with a low MOI of 0.01 or 0.001, mimicking the natural initial infection.  The levels of the chemokines CCL5 and CXCL10 in the basal media induced by the 3 strains were approximately equivalent at 24h p.i..  When the productivity per cell was calculated, the avian H5N1 strain induced higher secretion of CCL5 (RANTES) than the seasonal or pandemic strains.  Greater CCL5 secretion by H5N1 strains has been noted in in vitro NHBE cultures when a relatively high MOI is used compared to seasonal strains \cite{Chan2005, Chan2010, Zeng2011} and pandemic strains \cite{Zeng2011}.  MOI and degree of differentiation of NHBE are critical determinants of cyto/chemokine secretion \cite{Chan2010}.  However, in spite of the attenuated type 1 interferon response induced by H5N1 strains \cite{Zeng2007}, the chemokine secretion is not attenuated.
